Network Data Model and BerlinMOD Benchmark

### 2010-06-15 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
Network2 and TemporalNet2 added as students version for Mrs. Wrede.

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{2.9.StableNewFlob}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past several data models for the representation of histories of  spatio-temporal data
objects have been developed. Among others we can categorize these data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe the translation
from the \bmodb{} into the network constrained data model and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitudes.
%The very good results of the network data model in this comparison encourages us
%to propose a extension of the \bmodb{} with a additional query set BerlinMOD/Net.
%This new query set covers the specialised challenges for network constrained
%data models and enables us to use the \bmodb{} to compare the capabilities of
%different network data models with respect to this specialised network challenges.
\end{abstract}
\section{Introduction}
In the past several data models for the representation of spatial and
spatio-temporal data objects have been developed. Among others we can categorize
them into data models for objects moving freely in two dimensional space and
data models for network constrained moving objects. For both categories
several different data models have been presented like
\cite{chenzaniolosqlst,335426} for spatio-temporal data objects moving freely in
two dimensional space and \cite{1146465,956692,VazWolfNetMod} for
spatio-temporal data objects that are constrained by given networks,
to name just a few. Objects which are restricted to use existing networks, like
cars are restricted to use road networks, can be represented as moving point object
in both data
models, while objects, which are not restricted by a given network, like people,
can be represented as moving point object only in a data models of free movement
in space.

But if everything can be represented by a data model of free movement
space why do we spend time on network data models? Now, our experiments
presented in this paper show that the network data model outperforms the data model of
free movement in two dimensional space significantly. The network data model uses
less than 60\% of the storage space of the data model of free movement in space, mostly caused
by the fact that the number of units for the moving point objects in the network
data model is less than 50\% of the number of units in the data model of free
movement in two dimensional space. The total run time of the \bmodb{} for the
network data model is less than 50\% of the total run time of the data model of
free movement in two dimensional space. We think that this results show that
it is useful to develop specialised data models for specialised data structures
like network data models for network constrained moving objects to save storage
space and reduce query run times.

For this paper we choosed two representatives one for each data model category.
Both representatives are available in the \secondo{} DBMS and use the same
temporal representation. So we can exclude that different DBMS or temporal
representation issues bias the results of our data model comparison.

The network constrained data model is represented by
the network data model presented in \cite{1146465} and the data model of free
movement in two dimensional space is represented by the data model presented in
\cite{335426}.

For us it seems logical to use the \bmodb{} \cite{BerlinMODVLDB} which is also
available in the \secondo{} DBMS to compare the capabilities of the
both data models. The \bmodb{} is best to our knowledge the first benchmark for
complete spatio-temporal database systems and it is developed in the \secondo{}
DBMS. Furthermore the data generated by the \bmodb{} data generator is restricted
to the streets of the German capital Berlin, such that it can be translated into
a network constrained environment. And not at last the data model used in the
\bmodb{} is the data model of free movement in two dimensional space that we use
for our comparison. So we only have to translate the spatial and spatio-temporal
data types of the \bmodb{} once into our network data model representation.
This simplifies the control of the query results and avoids errors caused by
translation errors.

The translation of the spatial and spatio-temporal data types of the \bmodb{} data
into the network data model representation described in this paper can be seen as
example for the usage of the \bmodb{} with other compatible data representations or
DBMS.

%The very good results of the network data model encourages us to propose the extension
%of the \bmodb{} by a query set BerlinMOD/Net. The new query set covers the specialised
%challenges of network data models like network distance computing which are not
%covered by the original \bmodb{}. The new query set should enable us to compare
%the capabilities of different network data models with respect to the special
%challenges of network constrained data models.

The rest of the paper is organised as follows: In section \ref{sec:relWork} we give a
short reminder of the underlying \secondo{} DBMS (\ref{sec:secondo}), the \bmodb{}
(\ref{sec:bmodb}) and the two data models (\ref{sec:bmodbdatamod}
and \ref{sec:netdatamod}) we choose as representatives for our comparison.
The translation of the \bmodb{} data and query set in the network data model
representation is described in section \ref{sec:Translation}. The resulting
experimental benchmark setup is described in section \ref{sec:scenario} followed
by the results of our experiments in section \ref{sec:results}.
%At least we present our proposed extension of the \bmodb{} for network data models in
%section \ref{sec:newqueries}.
We conclude our work in section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
In the past other data models for free movement in the space
\cite{chenzaniolosqlst,335426} and for network constrained movement have been
presented \cite{1146465,956692,VazWolfNetMod}. As well there are some more
benchmarks \cite{COSTBenchmark, QueriesTheodoridis} and database systems for
spatial and spatio-temporal data types \cite{1054151,HERMES}. But best to our
knowledge they don't actually provide a combination of different implemented
and supported data models together with a existing benchmark feasible for moving
objects in space and network constrained moving objects like the actual
\secondo{} DBMS. So it is self-evident for us to use the \secondo{} DBMS in
combination with the provided data models and the \bmodb{} to compare the
capabilities of the network constrained data model and the data model of free
movement in two dimensional space provided with the \secondo{} DBMS.

In the next subsections we give short reminders of the \secondo{} database system
(\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}), and the both data models
(\ref{sec:bmodbdatamod}, \ref{sec:netdatamod}) we used in our experiments.
\subsection{Secondo}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{686903,1054151} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce  new data types and operations on
this data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the user may publish his extensions
as \secondo{} plugin such that other users can use this plugin to extend their own
\secondo{} system to use the new provided functionalities or repeat the published experiments.
\secondo{} is free available in the web \cite{secondoweb} and comes with a number
of already implemented spatial and spatio-temporal data types
and operations including the spatio-temporal data model of free movement in the
two dimensional space \ref{sec:bmodbdatamod} and the network data model
\ref{sec:netdatamod} used in this paper. Furthermore the \bmodb{} described in
\ref{sec:bmodb} has been developed in the \secondo{} DBMS. For our experiments
we used the \secondo{} version \secver{}.
\subsection{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was presented in \cite{BerlinMODVLDB} \nocite{BerlinMOD} and the
provided scripts for the data generator are implemented as \secondo{} DBMS operations.
The \bmodb{} is available in the web \cite{berlinmodweb} and provides a well defined
data-set and queries for the experimental evaluation of the capabilities of
spatial and spatio-temporal database systems dealing with histories of moving
objects. The \bmodb{} emphasises the development of complete systems
and simplifies experimental repeatability pointing out the weakness and the potency's
of the benchmarked systems.

The data-sets of the \bmodb{} are created using the street map of the German
capital Berlin \cite{bbike} and statistical data about the regions of Berlin
\cite{bevberlin,berlinstadtatlas} as input relations.
The created moving objects represent cars driving in the streets of Berlin,
simulating the behaviour of people living and working in Berlin.
Every moving object has a home node and a work node. Every weekday each car will
do a trip from the home node to work the work node in the morning and vice versa
in the late afternoon. Beside this randomly chosen cars will make additional
trips in the evening and up to six times at the weekend to randomly chosen
targets in Berlin and back home. The \bmodb{} uses the data model of free movement in two
dimensional space described in section \ref{sec:bmodbdatamod}. Because the \bmodb{}
generates all data sets restricted to the street map of Berlin the \bmodb{} can
also be used for network constrained data models, if the spatial and spatio-temporal
data types are translated into a corresponding network data model, like we did
for our experiments.

The number of observed cars and the duration of the observation period can be
influenced by the user setting the $scalefactor$ to different values in the data
generation script of the \bmodb{}. For example at $scalefactor$ 1.0 the data generator
creates 2000 moving point objects observed for 28 days. Each of them sending a
GPS-signal every 2 seconds. This simulated signals are simplified such that time
intervals when a car doesn't move or moves in the same direction at the same
speed are merged into one single time interval. For example: If a car parks in front
of the work node for 8 hours there will be only one entry in the history of the
cars movement with a time interval of 8 hours instead of 14.400 entries
one for each GPS time interval.

The \bmodb{} provides two different approaches to store the histories of moving
objects. On the one hand the object-based approach (OBA) and on the other hand
the trip based approach (TBA).

In the OBA the complete history for each moving object is kept together into
one single entry. There is only one relation $dataScar$
containing one tuple for each object consisting of the spatio-temporal data of
the object $journey$, the $licence$, the $type$, and the $model$ of the object.

In the TBA we have two relations $dataMcar$ and $dataMtrip$. $dataMcar$ contains
the static data for each object like $licence$, $type$, and $model$ together with
an object identifier $moid$. $dataMtrip$ contains for each $moid$ several tuple
each of them containing all units of a single trip of the moving object or a
single unit for a longer stop. For example each time the car drives from home node
to work node is a single trip and each time the car parks in front of the office
is also a single trip.

Beside the moving point objects the \bmodb{} provides several data sets each of them
containing 100 pseudo randomly generated data objects which are used in the
benchmark queries. Table \ref{tab:queryobjects} gives an overview of this query
objects. The \bmodb{} deals also with subsets from this query object sets consisting
of the first or second 10 query objects of a query object set. They are labeled
by the name of the query object set followed by a 1 for the first 10 or a 2 for
the second 10 query objects of the query object set.

\begin{table}[H]
  \begin{tabularx}{1.0\textwidth}{|l|X|}
    \hline
    \textbf{Name of Data Set}&\textbf{Tuple Content}\\
    \hline
    $QueryPoints$&Object identifier and \dt{point} values.\\
    \hline
    $QueryRegions$&Object identifier and \dt{region} value.\\
    \hline
    $QueryInstants$&Object identifier and time stamps.\\
    \hline
    $QueryPeriods$&Object identifier and space of time.\\
    \hline
    $QueryLicences$& Object identifier and a \dt{string} representing a licence value.\\
    \hline
  \end{tabularx}
  \caption{Query Object Relations of \bmodb{}}
  \label{tab:queryobjects}
\end{table}

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item known or unknown object identity,
  \item standard, spatial, temporal, or spatio-temporal dimension,
  \item point, range, or unbounded query interval,
  \item single object or object relations condition type,
  \item with or without aggregation.
\end{itemize}
We will present the 17 queries in more detail in section \ref{sec:queries}
together with our network data model algorithms for this queries.
\subsection{Data Model of Free Movement in two dimensional space}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the same data model of free movement in
two dimensional space presented in \cite{594784,335426,352963}. Spatial
positions are assumed to be located into a two dimensional space. A single spatial
position is represented by the data type \dt{point}. A \dt{point} consists of a pair of
\dt{real} values interpreted as x,y-coordinates in the assumed two dimensional plane.

Streets are represented by \dt{line} values. A \dt{line} value consists of a set
of $half segments$ representing the geometry of the line in the two dimensional space.
Each $half segment$ consists of two \dt{point} values and a Boolean flag telling us
if the left or the right point is the dominating point of the $half segment$.

Regions are represented by the data type \dt{region}. A \dt{region} consists of
a set of $half segments$ defining the outer border of the region in
the two dimensional space. If a region contains wholes the inner border is also
formed by the $half segments$.

In \secondo{} all this spatial data types and many standard data types can be
lifted to become time dependent \dt{moving} values. For all data types \dt{$\alpha$}
the constructor \dt{moving} creates a new data type \dt{moving}(\dt{$\alpha$})
(short form \dt{m$\alpha$}).

A car may be represented by a \dt{moving}(\dt{point}) short \dt{mpoint}.
A \dt{mpoint} consist of a set of units called
\dt{unit}(\dt{point}) (short form \dt{upoint}). Each \dt{upoint} consists of a time
interval and two \dt{point} values. The first \dt{point} value represents the
position of the \dt{mpoint} at the start of the time interval and the second
\dt{point} value represents the position of the \dt{mpoint} at the end of the
time interval. It is assumed that the object represented by the \dt{mpoint}
moves on the straight line between this two points with constant speed within the
given time interval. The velocity of the object is given by the ratio from the
distance of the two points and the length of the time interval of the unit.
All units of a \dt{mpoint} must have disjoint time intervals, because a car
cannot be at two different positions at the same time.
The units are sorted by ascending time intervals.

This spatio-temporal data model of \dt{moving} allows us to compute the position
of a \dt{mpoint} at every time instant within its definition time.
We can also compute the time instant the point passed a
given position assumed the \dt{mpoint} ever passes this position. The position of a
\dt{point} at a given time instant is represented by a \dt{intime}(\dt{point})
(short form \dt{ipoint}). A \dt{ipoint} consists of a time instant and a \dt{point} value.

Some other data types of \secondo{} which are used in the \bmodb{} are shown in
table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{1.0\textwidth}{|l|X|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type.\\
\hline
\dt{int} & Usual integer number.\\
\hline
\dt{real} & Usual real number.\\
\hline
\dt{instant} & A point in time.\\
\hline
\dt{periods} & A set of disjoint and not connected spaces of time.\\
\hline
\dt{mbool} & A time dependent boolean value, which will be constant \true{} or \false{}
within each \dt{ubool} \\
\hline
\dt{mreal} & Time dependent real number. Each unit will be defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\subsection{Network Data Model}
\label{sec:netdatamod}
The central idea of the network data model presented in \cite{1146465} is that
every movement is constrained by a given network and every position can be described
relative to this network. The data type \dt{network} is the central
data type in the network data model. All other data types of the network data model
are related to a given \dt{network} object by the unique network identifier that
is part of each \dt{network} object.

@@

The $sections$ relation of the \dt{network} object contains the
attributes of the street parts between two crossings or a crossing and
the end of the street. This are the route identifier of the street the section
belongs to, the tuple identifier of this street in the routes relation, the start
position and end position of the section on the street, the section curve, and
again two Boolean flags with the same meaning as in the routes relation.

@@

We created four B-Tree indexes for the route identifier attributes in the $routes$,
$junctions$ and $sections$ relation, and a R-Tree index over the curve attribute
of the $routes$ relation to support faster execution of operations dealing with
the relations of the \dt{network} object.

The data type \dt{gpoint} represents single positions in a given network. Besides the
network identifier a \dt{gpoint} consists of a route identifier, a distance from
the start of the route to the position of the \dt{gpoint} and a $side$ value
({$up$, $down$, $none$) telling us if the position is reachable from the $up$
or the $down$ side of the route in case of separated lanes. For simple
streets or positions which are reachable from both sides of the route the side
value is always $none$.

Parts of the network, regardless if they represent paths or regions, are given as
\dt{gline} values. Besides the network identifier a \dt{gline} consists of a set
of $route interval$s, and two Boolean flags. The Boolean flags tell us if the
\dt{gline} is defined and if the set of $route interval$s is sorted.

Each $route interval$ consists of a route identifier identifying the route
the route interval belongs to, and the start and the end position from the
route interval on this route
\footnote{In the original paper the $route interval$ includes a $side$ value
analogous to the \dt{gpoint}. But this parameter is not part of the implementation yet.}.
We call a \label{sec:sortedgline} set of $route interval$s sorted if the following
conditions are fullfilled:
\begin{itemize}
	\item all $route interval$s are disjoint
	\item the $route interval$s are stored in ascending order of their route identifiers
	\item if two disjoint $route interval$s have the same route identifier the
$route interval$ with the smaller start position is stored first
	\item for all $route interval$s the $start position \le end position$
\end{itemize}
Many algorithms take profit from sorted \dt{gline} values. For example: If $n$
is the number of the route intervals in a \dt{gline} the
decision, if a \dt{gpoint} is inside the \dt{gline} needs O($n$) time for unsorted
and O($\log n$) time for sorted \dt{gline} values,

@@

Query 9 asks for the longest distance travelled by a single vehicle during each
of the periods from $QueryPeriods1$. We restrict all trips to the periods, compute
the driven distances and select the maximum length for each query periods value.
Again we have to do a additional aggregation of the distances driven from the same
car in the same period in the TBA.

@@

of the \bmodb{} queries and the network data model queries for the object based (OBA)
and the trip based approach (TBA). We detected that the source data of the street

>>>>

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{2.9.x}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, and the translation
from the \bmodb{} into the network constrained data model and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space and data models
for network constrained objects. For both categories several diffferent data models
have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis,HERMESMDCPelekis}
for spatio-temporal data objects moving freely in two dimensional space (DMFS) and
\cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,NetworkVazirgiannis}
for spatio-temporal data objects that are constrained by given networks (NCDM),
to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

Why do we spend time on NCDM,if everything can be represented by
a DMFS? Now, it is natural to give positions related to the street
network instead of x,y-coordinates. NCDM are expected to use less storage space,
because geographical informations about street curves are stored only once in
the network, whereas in DMFS each street curve is stored in each moving point
object using this street. NCDM can support query processing with spezialized indexes,
because of the knowledge of the underlying network. And it is much easier to
formulate queries about the relationships between moving objects and the network
in the NCDM. And last but not least the results of our experiments show that
our network data model outperforms our data model of free movement in two dimensional
space by orders of magnitude. The network constrained data model uses less than
60\% of the storage space and less than 50\% of the total query run time of the
data model of free movement in space we used in our experiments.
We think that these results show that it is useful
to develop specialised data models for specialized data structures like NCDM for
newtork constrained moving objects to save storage space and reduce query run times.

For our benchmark experiments, presented in this paper, we chosed two data models
one for each data model category. Both data models use the same temporal representation
and are available in \secondo{} DBMS
\cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting}.
So we can exclude that different DBMS or temporal
representation issues bias the results of our data model comparison with the
\bmodb{}\cite{BerlinMODVLDBDuentgen}.
The DMFS we use is the data model presented in \cite{DataModelDataStructureGueting}
(SPACE for short).
And the NCDM we use is the data model presented in \cite{NetworkGueting} (NET for short).
Our implementation of NET is described first in this paper (Section \ref{sec:implNDM}
and provides some further concepts which were only sketched in \cite{NetworkGueting}.

We used the \bmodb{} \cite{BerlinMODVLDBDuentgen} to compare the capabilities of
the two data models, because the \bmodb{} is to the best of our knowledge the
first benchmark for complete spatio-temporal database systems.
It is developed and available in \secondo{} DBMS. And the data generated by the
\bmodb{} data generator are restricted to the streets of the German capital Berlin,
such that they can be translated into a network constrained environment.
And last but not least, the data model used in the \bmodb{} is SPACE that we use
for our comparison. So we only have to translate the spatial and spatio-temporal
data types of the \bmodb{} once into our NET representation. This simplifies the
control of the query results and avoids errors caused by translation.
The translation of the spatial and spatio-temporal data types of the \bmodb{} data
into the NET representation described in this paper \ref{sec:Translation} can be
seen as an example for the usage of the \bmodb{} with other compatible data
representations or DBMS.

The rest of the paper is organised as follows: Related Work is presented in Section
\ref{sec:relWork}. In the sections \ref{sec:secondo} - \ref{sec:bmodb} we give
short reviews of the underlying \secondo{} DBMS (Section \ref{sec:secondo}), the two data
models (SPACE Section \ref{sec:bmodbdatamod} and NET Section \ref{sec:netdatamod})
we choose for our comparison, and the \bmodb{} (Section \ref{sec:bmodb}).
In Section \ref{sec:implNDM} we give some information about our NET implementation,
the used operations and indexes.
The translation of the \bmodb{} data and query set into the NET representation
is described in Section \ref{sec:Translation}. The resulting experimental benchmark
setup is described in Section \ref{sec:scenario} followed by the results of our
experiments in Section \ref{sec:results}. We conclude our work in Section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
In the past many different spatio-temporal data models have been presented. Many
of them support only discrete spatio-temporal changes
\cite{sqlstchen,Ramadrachan,Langran1,Langran2,HunterWilliamson}
or deal only with current and future positions of continuously moving objects
\cite{MOSTWolfson}. In this paper we will focus on spatio-temporal data models
for complete histories of continuously moving objects.

To the best of our knowledge only a few of these data models have been completely
implemented into different database management systems:
\begin{enumerate}
  \item HERMES \cite{HERMESMDCPelekis, HERMESPelekis} is implemented as data
cartridge for the commercial Oracle\textregistered ORDBMS \cite{Oracle}. The
used data model is the DMFS STAU \cite{STAUPelekis}.
  \item The Dynamic Transportation Network \cite{DynamicTransportNetworkDing} is
a NCDM implemented in the open source database project PostgreSQL \cite{PostgreSQL}.
  \item In the freely available \secondo{} DBMS \cite{SecondoPlatformPrototypingGueting}
two spatio-temporal data models have been implemented.
\cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}
as DMFS and \cite{NetworkGueting} as NCDM.
\end{enumerate}

Because \secondo{} provides a data model for each of our catgories we use \secondo{}
and the provided data models. Although the data model of HERMES provides more
capabilities than SPACE.
Both data models use the basic idea of a sliced representation for moving objects
introduced by \cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}.
The implementation of the sliced representation in SPACE supports only linear
interpolation, whereas HERMES arc interpolation is also supported. Both data
models support time varying geometries, which change their positions and / or
extent in space and time dimension. HERMES supports discret and continuous
changes, whereas SPACE only supports continuously moving objects. HERMES uses
one spatial object for different spatial geometries, whereas SPACE
uses different spatial objects for the different spatial geometries.
The same holds for operations on topological relationships. SPACE
uses different operations, whereas HERMES only provides a single operator which
distinguishes between the different topological relationships via a ``mask'' parameter.

Although HERMES provides a greater flexibility we decided to use SPACE in our
experiments, because the \secondo{} DBMS provides besides
the SPACE data model a NCDM \cite{NetworkGueting} (see Section \ref{sec:netdatamod}),
and with the \bmodb{} \cite{BMODBVLDBDuentgen} (see \ref{sec:bmodb})
to the best of our knowledge the first benchmark for complete spatio-temporal
database systems.

Other existing benchmarks \cite{COSTBenchmarkJensen,QueriesTheodoridis}

like \cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting,STAUPelekis,HERMESMDCPelekis}
and data models dealing with network constrained moving objects (NCDM) like \cite{NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing}.

and \cite{NetworkGueting} as NCDM. Therefore we describe these data models
in the sections \ref{sec:bmodbdatamod} - \ref{sec:netdatamod} in more detail.

\cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting,NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing},As mentioned before we can categorize spatio-temporal data models into
data models dealing with current and future positions of moving objects,
data models dealing with complete histories of moving objects, and data models
dealing with past, current and future positions of moving objects. We will focus
on data models dealing with complete histories of moving objects, like \cite{STAUPelekis,HERMESMDCPelekis,RepresentingMovingObjectsGueting,DataModelDataStructureGueting,NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing}, to name just a few.

DMFS\cite{STAUPelekis,HERMESPelekis,RepresentingMovingObjectsGueting,DataModelDataStructureGueting,sqlstchen}

NCDM\cite{NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing}.

Also, there are some more
benchmarks \cite{COSTBenchmarkJensen,QueriesTheodoridis} and database systems for
spatio-temporal data types \cite{SecondoPlatformPrototypingGueting,HERMESPelekis}. But to the best of our
knowledge they do not provide a combination of different implemented
and supported data models together with an existing benchmark feasible for moving
objects in space and network constrained moving objects like the current \secondo{} DBMS.
So it is natural for us to use the \secondo{} DBMS in combination with the provided
data models and the \bmodb{} to compare the capabilities of the two data models.

In the next sections we give short reviews of the \secondo{} database system
(Section \ref{sec:secondo}), the both data models (Section \ref{sec:bmodbdatamod},
Section \ref{sec:netdatamod}) we used in our experiments, and the \bmodb{}
(Section \ref{sec:bmodb}).
\section{\secondo{} DBMS}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce new data types and operations on
these data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the users may publish their extensions
as a \secondo{} plugin such that other users can use this plugin to extend their own
\secondo{} system. They may use the newly provided functionalities or repeat the
published experiments. \secondo{} is freely available on the web \cite{secondoweb}.
It comes with a number of already implemented spatial and spatio-temporal data types
and operations including the DMFS (Section \ref{sec:bmodbdatamod}) and the NDM
(Section \ref{sec:netdatamod}) used in this paper. Furthermore, the \bmodb{}
described in Section \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version \secver{}.
\section{Data Model of Free Movement in Two Dimensional Space(SPACE)}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the DMFS presented in \cite{STDTModelingQueryingMOinDErwig,DataModelDataStructureGueting,RepresentingMovingObjectsGueting}.
Spatial positions are assumed to be located in a two dimensional space.

A single spatial
position is represented by the data type \dt{point}. A \dt{point} consists of a pair of
\dt{real} values interpreted as x,y-coordinates in the assumed two dimensional plane.

Streets are represented by \dt{line} values. A \dt{line} value consists of a set
of \dt{HalfSegment}s representing the geometry of the line in the two dimensional space.
Each \dt{HalfSegment} consists of two \dt{point} values and a Boolean flag telling us
if the left or the right point is the dominating point of the \dt{HalfSegment}.

Regions are represented by the data type \dt{region}. A \dt{region} consists of
a set of \dt{HalfSegment}s defining the outer border of the region in
the two dimensional space. If a region contains wholes the inner border is also
formed by the \dt{HalfSegment}s.

In \secondo{} all these spatial data types and many standard data types can be
''lifted'' to become time dependent \dt{moving} values. For all data types \dt{$\alpha$}
the constructor \dt{moving} creates a new data type \dt{moving}(\dt{$\alpha$})
(short form \dt{m$\alpha$}).

A car may be represented by a \dt{moving}(\dt{point}), short \dt{mpoint}.
An \dt{mpoint} consists of a set of units called
\dt{unit}(\dt{point}) (short form \dt{upoint}). Each \dt{upoint} consists of a time
interval and two \dt{point} values. The first \dt{point} value represents the
position of the \dt{mpoint} at the start of the time interval and the second
\dt{point} value represents the position of the \dt{mpoint} at the end of the
time interval. It is assumed that the object represented by the \dt{mpoint}
moves on the straight line between these two points with constant speed within the
given time interval. The velocity of the object is given by the ratio from the
distance of the two points and the length of the time interval of the unit.
All units of an \dt{mpoint} must have disjoint time intervals, because a car
cannot be at two different positions at the same time.
The units are sorted by ascending time intervals.

This spatio-temporal data model of \dt{moving} allows us to compute the position
of an \dt{mpoint} at every time instant within its definition time.
We can also compute the time instant the point passed a
given position assuming the \dt{mpoint} ever passes this position. The position of a
\dt{point} at a given time instant is represented by an \dt{intime}(\dt{point})
(short form \dt{ipoint}). An \dt{ipoint} consists of a time instant and a \dt{point} value.

Some other data types of \secondo{} which are used in the \bmodb{} are shown in
Table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{1.0\textwidth}{|l|X|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type\\
\hline
\dt{int} & Usual integer number\\
\hline
\dt{real} & Usual real number\\
\hline
\dt{instant} & A point in time\\
\hline
\dt{periods} & A set of disjoint and not connected time interval\\
\hline
\dt{mbool} & A time dependent boolean value, which is constant \true{} or \false{}
within each \dt{ubool} \\
\hline
\dt{mreal} & Time dependent real number. Each unit is defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\section{Network Data Model(NET)}
\label{sec:netdatamod}
The central idea of the NDM presented in \cite{NetworkGueting} is that
every movement is constrained by a given network and every position can be described
relative to this network. The data type \dt{network} is the central
data type in the NDM. All other data types of the NDM
are related to a given \dt{network} object by the unique network identifier that
is part of each \dt{network} object.

@@

The $sections$ relation of the \dt{network} object contains the
attributes of the street parts between two crossings or a crossing and
the end of the street. These are the route identifier of the street the section
belongs to, the tuple identifier of this street in the routes relation, the start
position and end position of the section on the street, the section curve, and
again two Boolean flags with the same meaning as in the routes relation.

@@

We created four B-Tree indexes for the route identifier attributes in the $routes$,
$junctions$ and $sections$ relation, and an R-Tree index over the curve attribute
of the $routes$ relation to support faster execution of operations dealing with
the relations of the \dt{network} object.

The data type \dt{gpoint} represents single positions in a given network. Besides the
network identifier a \dt{gpoint} consists of a route identifier, a distance from
the start of the route to the position of the \dt{gpoint} and a $side$ value
({$up$, $down$, $none$) telling us if the position is reachable from the $up$
or the $down$ side of the route in case of separated lanes. For simple
streets or positions which are reachable from both sides of the route, the side
value is always $none$.

Parts of the network, regardless if they represent paths or regions, are given as
\dt{gline} values. Besides the network identifier, a \dt{gline} consists of a set
of \dt{RouteInterval}s, and two Boolean flags. The Boolean flags tell us if the
\dt{gline} is defined and if the set of \dt{RouteInterval}s is sorted.

Each \dt{RouteInterval} consists of a route identifier identifying the route that
the route interval belongs to, and the start and the end position of the
route interval on this route
\footnote{In the original paper the \dt{RouteInterval} includes a $side$ value
analogous to the \dt{gpoint}. But this parameter is not yet part of the implementation.}.
We call a \label{sec:sortedgline} set of \dt{RouteInterval}s sorted if the following
conditions are fullfilled:
\begin{itemize}
   \item all \dt{RouteInterval}s are disjoint
   \item the \dt{RouteInterval}s are stored in ascending order of their route identifiers
   \item if two disjoint \dt{RouteInterval}s have the same route identifier, the
\dt{RouteInterval} with the smaller start position is stored first
   \item for all \dt{RouteInterval}s, the $start$ $position \le end$ $position$
\end{itemize}
Many algorithms take profit from sorted \dt{gline} values. For example: If $n$
is the number of the route intervals in a \dt{gline}, the
decision, if a \dt{gpoint} is inside the \dt{gline} needs O($n$) time for unsorted
and O($\log n$) time for sorted \dt{gline} values.

@@

Query 9 asks for the longest distance travelled by a single vehicle during each
of the periods from $QueryPeriods1$. We restrict all trips to the periods, compute
the driven distances and select the maximum length for each query periods value.
Again we have to do an additional aggregation of the distances driven from the same
car in the same period in the TBA.

@@

of the \bmodb{} queries and the NDM queries for the OBA and the TBA. We detected
that the source data of the street map of the \bmodb{} is not well defined in all places.
